---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, options(readr.show_col_types = FALSE))
```

# Prepare datasets 

This document details the process of importing VirScan results and metadata for the ENDIA and VIGR cohorts, preparing the datasets by modifying relevant columns, and merging them for each cohort to enable downstream analysis.

```{r}
library(tidyverse)
```

### ENDIA

1. Read in and join VirScan data and `filter` out peptides and samples that are all 0.
2. Read in metadata 
3. Join VirScan data to metadata and export dataset to `cache`

```{r}
if(file.exists("cache/endia_virscan_metadata.rds")){
  endia_virscan_metadata <- read_rds("cache/endia_virscan_metadata.rds")
} else {
  #1.
endia_virscan_hits_counts_annot1_3 <- read_tsv("raw_data/virscan_hits_counts/phip11_plate1-3_v_kiwook_2_CDIVirScan_000_Hits_counts_annotated.tsv") %>% 
  select(-contains("Beads_Only")) %>% 
  pivot_longer(cols = starts_with("KiWook"), names_to = "participant_id", values_to = "abundance") 

endia_virscan_hits_counts_annot4_6 <- read_tsv("raw_data/virscan_hits_counts/phip11_plate4-6_v_kiwook_2_CDIVirScan_000_Hits_counts_annotated.tsv") %>% 
   select(-contains("Beads_Only"),
          -KiWook_PhIP11_Plate6_81_G9_POWH_SAMPLE) %>% #remove positive control 
  pivot_longer(cols = starts_with("KiWook"), names_to = "participant_id", values_to = "abundance")

endia_virscan_long <- bind_rows(endia_virscan_hits_counts_annot1_3, endia_virscan_hits_counts_annot4_6) %>% 
  group_by(pep_id) %>%
  filter(any(abundance > 0)) %>% # remove peptides that are 0 in all samples
  ungroup() %>% 
  group_by(participant_id) %>% 
  filter(sum(abundance) != 0) %>% # remove samples with 0 peptides detected 
  ungroup() %>% 
  mutate(sample_id = str_remove(participant_id, "^(?:[^_]+_){5}"),   # keep everything after the 5th _
          timepoint = str_extract(sample_id, "[^_]+$"),           # extract last segment from sample_id
          participant_id = str_remove(sample_id, "_[^_]+$"),        # remove timepoint from sample_id
          .keep = "unused")
  #2.

endia_samples <- read_csv("raw_data/metadata/identify_plasma_samples.csv") %>% 
  mutate(participant_id = str_replace_all(structured_participant_id, "-", "_"), .keep = "unused") 
         
endia_metadata <- readxl::read_xlsx("raw_data/metadata/finalweights_teddy_plasma_with_visits_deidentified_confounders.xlsx") %>% 
  right_join(endia_samples, by = join_by(mother_id, infant_id)) %>% 
  relocate(participant_id) %>% 
  mutate(recorded_visit = str_replace(recorded_visit, "(?<=[BVT])0", ""), # replace 0 preceded by B, V or T
         condition = ifelse(case == 1, "Case", "Control")) %>% 
  unite(sample_id, c(participant_id, recorded_visit), sep = "_", remove = FALSE) %>% 
  distinct(sample_id, .keep_all = TRUE) %>% # remove duplicate sample IDs from the NCC methodology
  group_by(participant_id) %>% # remove controls that seroconverted to cases
  filter(!(all(c("Case", "Control") %in% condition) & condition == "Control")) %>% 
  ungroup()

  #3.

endia_virscan_metadata <- endia_virscan_long %>% 
  left_join(endia_metadata, by = join_by(sample_id, participant_id)) 

write_rds(endia_virscan_metadata, "cache/endia_virscan_metadata.rds")

}
```

### VIGR

For VIGR we similarly read in the metadata files and VirScan data and join all datasets together.
We also filter out peptides that are all 0. Unlike in ENDIA, no samples had a total abundance of 0, so none needed to be removed. 
A column called `maternal_T1D` was created to indicate the T1D status of the mother to match ENDIA for statistical analyses.

```{r}
if(file.exists("cache/vigr_virscan_metadata.rds")){
  vigr_virscan_metadata <- read_rds("cache/vigr_virscan_metadata.rds")
} else {
  
vigr_extra_metadata <- readxl::read_xlsx("raw_data/metadata/VIGR_HLA.xlsx") %>%
  select(Participant_Id, UNSW, Diabetes, Risk_group, `Age (years)`) %>% 
  mutate(maternal_T1D = if_else(Diabetes == "Mother", "T1D", "no T1D"))

vigr_metadata <- read.csv("raw_data/metadata/vigr_sample_info.csv")[-1] %>% 
   mutate(Replicate = paste0("t", Replicate)) %>% 
   left_join(vigr_extra_metadata, join_by(Participant_Id, UNSW)) %>% 
   rename(sample_id = Participant_Id)

vigr_virscan_metadata <- read_tsv("raw_data/virscan_hits_counts/vigr_Hits_counts_annotated.tsv") %>% 
  pivot_longer(cols = starts_with("UNSW"), names_to = "participant_id", values_to = "abundance") %>%
  select(!starts_with("Beads")) %>%
  group_by(pep_id) %>% 
  filter(any(abundance > 0 )) %>% 
  ungroup() %>% 
  left_join(vigr_metadata, by = join_by("participant_id" == "Sample")) %>% 
  filter(Replicate != "t1") #remove cases and ass. controls before seroconversion

write_rds(vigr_virscan_metadata, "cache/vigr_virscan_metadata.rds")
}
```


