---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, options(readr.show_col_types = FALSE))
```

# Generate figures using the AVARDA output from the ENDIA cohort examining cross reactivity and indistinguishable viruses

This replicates:

- Supplementary Fig. 4 | Cross-reactivity of detected EV peptides by AVARDA
- Supplementary Data Fig. 5 | Viral species classified as indistinguishable by AVARDA

```{r}
library(tidyverse)
library(UpSetR)
```

Read in ENDIA VirScan with metadata dataset combined in `00_prepare_datasets.Rmd`
and `filter` to only keep ENDIA onset samples (after seroconversion)
```{r}
endia_virscan_onset <- read_rds("cache/endia_virscan_metadata.rds") %>% 
   filter(onset_visit == 1)
```

Read in and combine AVARDA results for ENDIA and only retain the onset samples
```{r}
avarda_1 <- read.csv("raw_data/avarda/phip11_plate1-3_v_kiwook_2_AVARDA_compiled_full_output.csv", header = TRUE)
avarda_2 <- read.csv("raw_data/avarda/phip11_plate4-6_v_kiwook_2_AVARDA_compiled_full_output.csv", header = TRUE)

endia_avarda_onset <- avarda_1 %>% 
  bind_rows(avarda_2) %>% 
  filter(!str_detect(name, "SAMPLE")) %>% #remove positive control
  separate_wider_delim(name, delim = "_", names = c(NA, NA, NA, NA, NA, "col", "to", "merge", "too", "timepoint")) %>% 
  unite(name, col:too) %>% 
  unite(sample_id, name, timepoint, remove = FALSE) %>% 
  mutate(genus = str_extract(Virus, "(?<=g__)[^.]+"),
         species = str_match(Virus, ".*__(.*)$")[, 2]) %>% 
  filter(sample_id %in% endia_virscan_onset$sample_id)

```

### UpSet plot to show cross reactivity of EV peptides between classified *Enterovirus* species

Obtain classified *Enterovirus* peptides 

```{r}
endia_onset_ev_peptides <- endia_avarda_onset %>% 
  filter(genus == "Enterovirus") %>% 
  unite(peptides, Evidence.Peptides, XR.peptides, sep = "|") %>%
  separate_longer_delim(peptides, delim = "|") %>% 
  filter(species != "Human_enterovirus") %>% 
  mutate(species = str_replace_all(species, "_", " "))

endia_ev_species_pep_list <- endia_onset_ev_peptides %>% 
 group_by(species) %>% 
    summarise(peptides = list(unique(peptides))) %>% 
    deframe()
```

Plot the UpSet plot 

```{r, fig.height=7, fig.width=8}
endia_ev_upset_plot <- upset(
  fromList(endia_ev_species_pep_list),
  order.by = "freq",
  nsets = 8,
  main.bar.color = "#0072B2",
  sets.x.label = "No. peptides per species",
  mainbar.y.label = "Shared VirScan peptides",
  queries = list(
    list(
      query = intersects,
      params = list("Rhinovirus A", "Rhinovirus B", "Enterovirus B", "Enterovirus C", 
                    "Enterovirus D", "Enterovirus A", "Enterovirus H", "Rhinovirus C"),
      color = "#CC79A7",
      active = TRUE )))

endia_ev_upset_plot
```

**Supplementary Fig. 4:** Cross-reactivity of detected EV peptides by AVARDA
UpSet plot illustrating all *Enterovirus* genus peptides detected in the ENDIA dataset after being processed using the AVARDA algorithm, and their assignment to specific viral species. Each vertical bar (top) represents the total number of *Enterovirus* peptides shared by each species, ordered by decreasing count. The pink bar highlights the number of peptides shared across all eight Enterovirus species. The black horizontal bars (left) indicate the total number of peptides assigned to each individual *Enterovirus* species. Only classified *Enterovirus* species were included.

```{r, echo = FALSE, eval = FALSE}
pdf("figures/endia_ev_upset_plot.pdf")
endia_ev_upset_plot
dev.off()
```

### Indistinguishability of species

```{r, fig.width=9}
endia_indis_plot <- endia_avarda_onset %>%
  mutate(indis_tag = ifelse(Indistinguishablity.Groups >= 1, 1, 0),
         species = str_replace_all(species, "_", " ")) %>%
  group_by(genus, species) %>%
  summarise(indis_samples = sum(indis_tag), .groups = "drop") %>%
  filter(indis_samples > 5) %>%
  ggplot(aes(x = indis_samples, y = reorder(species, indis_samples))) +
  geom_bar(aes(fill = genus), stat = "identity") +
  scale_fill_manual(values = c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#e6ab02")) +
  labs(x = "No. of samples", y = "Virus species") +
  theme_minimal(base_size = 12) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 10, face = "italic"))

endia_indis_plot
```

```{r, echo=FALSE}
ggsave("figures/endia_indis_plot.png", endia_indis_plot, width = 9, height = 7, dpi = 350)
```

**Supplementary Fig. 5:** Viral species classified as indistinguishable by AVARDA in the ENDIA cohort. 
Bar plot depicting the number of ENDIA samples in which each viral species was classified as indistinguishable from another species by the AVARDA algorithm. Bars are ordered by number of samples and are colour-coded by viral genus. Species from the *Enterovirus* and *Mastadenovirus* genera were most frequently indistinguishable from others across the ENDIA cohort. 
