---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, options(readr.show_col_types = FALSE))
```

# Statistical testing of antigen levels against motifs in *Enterovirus* VP1 and 3D region peaks

```{r}
library(tidyverse)
source("scripts/read_blast.R")
library(glmmTMB)
library(gt)
```

Use data prepared in `01_figure_01_CXVB_antigen_mapping.Rmd`

```{r}
endia_virscan_onset <- read_rds("cache/endia_virscan_metadata.rds") %>%
  filter(onset_visit == 1) %>% 
  group_by(sample_id) %>% 
  mutate(rpk = abundance / sum(abundance) * 100000) %>% 
  mutate(log_lib = log(sum(abundance) + 1e-8)) %>% # add offset for testing 
  ungroup()

vigr_virscan_metadata <- read_rds("cache/vigr_virscan_metadata.rds") %>% 
  group_by(sample_id) %>% 
  mutate(rpk = abundance / sum(abundance) * 100000) %>% 
  ungroup()
```

```bash
makeblastdb -in raw_data/conserved_ev_motifs.fasta -dbtype prot -out raw_data/blast_databases/conserved_ev_motifs_db

blastp -task blastp-short -query cache/endia_virscan_hits_peptides.fasta -db raw_data/blast_databases/conserved_ev_motifs_db -outfmt '6 qaccver saccver pident nident length evalue bitscore mismatch gapopen qstart qend sstart send qseq sseq ppos stitle frames' -evalue 0.01 -word_size 2 -out raw_data/blast_results/blastp_endia_ev_motifs.blast

blastp -task blastp-short -query cache/vigr_virscan_hits_peptides.fasta -db raw_data/blast_databases/conserved_ev_motifs_db -outfmt '6 qaccver saccver pident nident length evalue bitscore mismatch gapopen qstart qend sstart send qseq sseq ppos stitle frames' -evalue 0.01 -word_size 2 -out raw_data/blast_results/blastp_vigr_ev_motifs.blast
```

```{r}
endia_blastp_ev_motifs <- read_blast("raw_data/blast_results/blastp_endia_ev_motifs.blast")

vigr_blastp_ev_motifs <- read_blast("raw_data/blast_results/blastp_vigr_ev_motifs.blast")
```

```{r}
endia_motif_results <- endia_blastp_ev_motifs %>%
    left_join(endia_virscan_onset, by = join_by(qaccver == pep_id)) 

vigr_motif_results <- vigr_blastp_ev_motifs %>% 
   left_join(vigr_virscan_metadata, by = join_by(qaccver == pep_id))
```

## ENDIA

```{r}
endia_motif_w_genera <- endia_motif_results %>%
  group_by(saccver, condition, taxon_genus) %>% 
  summarise(total_rpk = sum(rpk),
            no_of_peptides = n_distinct(qaccver),
            no_of_participants = n_distinct(participant_id),
            no_of_samples = n_distinct(sample_id),
            .groups = "drop") %>% 
  filter(total_rpk > 0)
```

```{r, fig.height=8}
endia_motif_w_genera %>%
  filter(taxon_genus == "Enterovirus") %>% 
  ggplot(aes(x = taxon_genus, y = total_rpk, fill = condition)) +
  geom_col(position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c("Case" = "#d73027", "Control" = "#4575b4"), labels = c("Case", "Control")) +
  geom_text(aes(label = no_of_peptides), 
            position = position_dodge(width = 0.9), 
            vjust = -0.2, hjust = -0.7, size = 3) +
  geom_text(aes(label = no_of_samples), 
            position = position_dodge(width = 0.9), 
            vjust = -2, hjust = -0.7, size = 3, color = "purple") +
  facet_wrap(~ saccver) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom",
        line = element_blank(),
      axis.text.x = element_blank(), 
      strip.text = element_text(face = "bold")) +
  labs(x = "Enterovirus peptides with significant hits", y = "Total normalised abundance", fill = "")
```

This essentially shows that the same peptides (black numbers) are present in all samples (purple number) between case and control

```{r, echo = FALSE, fig.height=8}
ggsave("figures/ev_hits_to_motifs.png", width = 8, height = 7)
```

Having a look at the BLAST results for the VP1 motif in ENDIA, stratified by sex:
Female controls appear to have a stronger reaction compared to male controls. However, male cases have a stronger reaction compared to female cases. Black numbers are number of peptides and purple number are number of samples

```{r, echo = FALSE}
endia_motif_results %>%
  group_by(saccver, condition, taxon_genus, infant_sex) %>% 
  summarise(total_rpk = sum(rpk),
            no_of_peptides = n_distinct(qaccver),
            no_of_participants = n_distinct(participant_id),
            no_of_samples = n_distinct(sample_id),
            .groups = "drop") %>% 
  filter(total_rpk > 0) %>% 
  filter(taxon_genus == "Enterovirus") %>% 
  filter(saccver == "VP1_motif") %>% 
  ggplot(aes(x = taxon_genus, y = total_rpk, fill = condition)) +
  geom_col(position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c("Case" = "#d73027", "Control" = "#4575b4"), labels = c("Case", "Control")) +
  geom_text(aes(label = no_of_peptides), 
            position = position_dodge(width = 0.9), 
            vjust = -0.2, hjust = -0.7, size = 3) +
  geom_text(aes(label = no_of_samples), 
            position = position_dodge(width = 0.9), 
            vjust = -2, hjust = -0.7, size = 3, color = "purple") +
  facet_wrap(~ infant_sex) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom",
        line = element_blank(),
      axis.text.x = element_blank(), 
      strip.text = element_text(face = "bold")) +
  labs(x = "Enterovirus peptides with significant hits to VP1 motif", y = "Total normalised abundance", fill = "")
```

```{r}
tidy_looksie <- function(glm_model) {
    glm_model %>% 
    broom.mixed::tidy(effect = "fixed", conf.int = TRUE, exponentiate = TRUE) %>% 
    select(term, estimate, p.value, conf.low, conf.high) %>% 
    mutate(significant = ifelse(p.value <= 0.10, T, F))}
```

### Stats ENDIA

```{r}
endia_motif_for_stats <- endia_motif_results %>% 
  filter(taxon_genus == "Enterovirus") %>% 
  group_by(saccver, sample_id) %>% 
  mutate(total_rpk = sum(rpk)) %>% 
  mutate(total_abundance = sum(abundance)) %>% 
  ungroup() %>% 
  select(saccver, sample_id, case, condition, taxon_genus, total_rpk, infant_id, mother_id, 
         deidentified_nest_id_new, age_at_sample_collection_days, infant_sex,
         infant_HLA, weightTEDDY, maternal_T1D, total_abundance, log_lib) %>% 
  distinct(saccver, taxon_genus, sample_id, .keep_all = TRUE) %>% 
  mutate(age_sample_collection_month = age_at_sample_collection_days / 365 * 12) %>% 
  mutate(log_total_rpk = log(total_rpk + 1)) %>% 
  mutate(across(c(infant_HLA, maternal_T1D, infant_id, mother_id, deidentified_nest_id_new, infant_sex), as.factor)) %>% 
  mutate(age_c = scale(age_at_sample_collection_days)) %>% 
  mutate(total_rpk_int = as.integer(total_rpk)) %>%  # also testing
  mutate(Condition = factor(condition, levels = c("Control", "Case"), labels = c("Control", "Case"))) # for testing purposes
```

#### ENDIA 3D region 

```{r}
endia_3d_motif_for_stats <- endia_motif_for_stats %>% filter(saccver == "3D_motif")

endia_3d_model <- glmmTMB(formula = case ~ log_total_rpk + infant_HLA + age_sample_collection_month + infant_sex + maternal_T1D + (1|deidentified_nest_id_new) + (1|mother_id),
       family = poisson(link = "log"),
        weights = weightTEDDY,
       data = endia_3d_motif_for_stats)

endia_3d_model %>% tidy_looksie()
```

##### 3D stratified by sex 

No significance in abundances but HLA is significant in females 

```{r}
endia_3d_motif_for_stats_xy <- endia_3d_motif_for_stats %>% filter(infant_sex == "Male")
endia_3d_motif_for_stats_xx <- endia_3d_motif_for_stats %>% filter(infant_sex == "Female")

endia_3d_model_xy <- glmmTMB(formula = case ~ log_total_rpk + infant_HLA + age_sample_collection_month + maternal_T1D + (1|deidentified_nest_id_new) + (1|mother_id),
       family = poisson(link = "log"),
        weights = weightTEDDY,
       data = endia_3d_motif_for_stats_xy)

endia_3d_model_xx <- glmmTMB(formula = case ~ log_total_rpk + infant_HLA + age_sample_collection_month + maternal_T1D + (1|deidentified_nest_id_new) + (1|mother_id),
       family = poisson(link = "log"),
        weights = weightTEDDY,
       data = endia_3d_motif_for_stats_xx)

endia_3d_model_xx %>% tidy_looksie()

endia_3d_model_xy %>% tidy_looksie()
```

3D doesn't seem to be a good predictor for IA. However, there is enrichment of 3D in cases 

This enrichment appears in a statistical test too:

```{r}
endia_3d_abundance_as_outcome <- glmmTMB(formula = total_abundance ~ Condition + infant_HLA + age_sample_collection_month + infant_sex + maternal_T1D + offset(log_lib) + (1|deidentified_nest_id_new) + (1|mother_id),
       family = poisson(link = "log"),
        weights = weightTEDDY,
       data = endia_3d_motif_for_stats) 

endia_3d_abundance_as_outcome %>% tidy_looksie()
```

#### VP1 

No significance in abundance but HLA is significant 

```{r}
endia_vp1_motif_for_stats <- endia_motif_for_stats %>% filter(saccver == "VP1_motif")

endia_vp1_model <- glmmTMB(formula = case ~ log_total_rpk + infant_HLA + age_sample_collection_month + infant_sex + maternal_T1D + (1|deidentified_nest_id_new) + (1|mother_id),
       family = poisson(link = "log"),
        weights = weightTEDDY,
       data = endia_vp1_motif_for_stats)

endia_vp1_model %>% tidy_looksie()
```
##### VP1 stratified by sex ENDIA

Stratifying by sex results in significant difference in abundance for males, but not for females. 
However, HLA is significant in females, but not for males.

Number of males and females per case or control group:
```{r}
endia_vp1_motif_for_stats %>% 
  count(condition, infant_sex, name = "count") %>% 
  arrange(infant_sex)
```

```{r}
endia_vp1_motif_for_stats_xy <- endia_vp1_motif_for_stats %>% filter(infant_sex == "Male")
endia_vp1_motif_for_stats_xx <- endia_vp1_motif_for_stats %>% filter(infant_sex == "Female")

endia_vp1_xy_model <- glmmTMB(formula = case ~ log_total_rpk + infant_HLA + age_sample_collection_month + maternal_T1D +  (1|deidentified_nest_id_new) + (1|mother_id),
       family = poisson(link = "log"),
        weights = weightTEDDY,
       data = endia_vp1_motif_for_stats_xy)

endia_vp1_xx_model <- glmmTMB(formula = case ~ log_total_rpk + infant_HLA + age_sample_collection_month + maternal_T1D + (1|deidentified_nest_id_new) + (1|mother_id),
       family = poisson(link = "log"),
        weights = weightTEDDY,
       data = endia_vp1_motif_for_stats_xx)

summary(endia_vp1_xy_model)
summary(endia_vp1_xx_model)
```

Pretty table with VP1 sex stratification results for ENDIA.
Sample count for males and females per case and control group are shown in "Number of males and females per case or control group" ^

```{r, echo = FALSE}
endia_vp1_xx_model_tidy <- endia_vp1_xx_model %>% 
  broom.mixed::tidy(effect = "fixed", conf.int = TRUE, exponentiate = TRUE) %>% 
    add_column(sex = "Females")

endia_vp1_xy_model_tidy <- endia_vp1_xy_model %>% 
  broom.mixed::tidy(effect = "fixed", conf.int = TRUE, exponentiate = TRUE) %>% 
    add_column(sex = "Males")

endia_vp1_sex_s_models_tidy <- rbind(endia_vp1_xx_model_tidy, endia_vp1_xy_model_tidy) %>% 
    dplyr::rename(RR = estimate,
               `CI lower` = conf.low,
               `CI upper` = conf.high) %>% 
    select(term, RR, `CI lower`, `CI upper`, p.value, sex) %>% 
    filter(term != "(Intercept)")

endia_vp1_sex_s_models_tidy_table <- endia_vp1_sex_s_models_tidy %>%
  mutate(term = case_match(term, !!!c("log_total_rpk" ~ "log(EV)"), .default = term)) %>% 
  select(-sex) %>% 
  gt(rowname_col = "term") %>%
  tab_header(
    title = "Summary of GLM model tests"
  ) %>%
  cols_label(
    term = "Predictor",
    RR = "RR",
    `CI lower` = "CI Lower",
    `CI upper` = "CI Upper",
    p.value = "Pr(>|z|)"
  ) %>%
  fmt_number(
    columns = c(RR, `CI lower`, `CI upper`, p.value),
    decimals = 3
  ) %>%
  cols_align(
    align = "left",
    columns = everything()
  ) %>% 
  tab_row_group(
    label = "ENDIA females",
    rows = 1:6
  ) %>% 
  tab_row_group(
    label = "ENDIA males",
    rows = 7:12
  ) %>% 
  tab_style(
    style = cell_fill(color = "lightgrey"),
    locations = cells_row_groups(groups = c("ENDIA males", "ENDIA females"))
  )

as_raw_html(endia_vp1_sex_s_models_tidy_table)

```

```{r, echo = FALSE}
gtsave(endia_vp1_sex_s_models_tidy_table, "figures/endia_vp1_sex_s_models_tidy_table.png")
```

No significance for sex in an interaction model (log_total_rpk * infant_sex)
```{r}
endia_vp1_model_sex_inter <- glmmTMB(formula = case ~ log_total_rpk * infant_sex + infant_HLA + age_sample_collection_month + maternal_T1D + (1|deidentified_nest_id_new) + (1|mother_id),
       family = poisson(link = "log"),
        weights = weightTEDDY,
       data = endia_vp1_motif_for_stats)

endia_vp1_model_sex_inter %>% tidy_looksie()
```

## VIGR

```{r}
vigr_motif_w_genera <- vigr_motif_results %>%
  group_by(saccver, Condition, taxon_genus) %>% 
  summarise(total_rpk = sum(rpk),
            no_of_peptides = n_distinct(qaccver),
            no_of_participants = n_distinct(participant_id),
            no_of_samples = n_distinct(sample_id),
            .groups = "drop") %>% 
  filter(total_rpk > 0)
```

Examining BLAST results shows that in VIGR, the same peptides (black numbers) are present in all participants (purple number) between case and control

```{r, fig.height=8}
vigr_motif_w_genera %>%
  filter(taxon_genus == "Enterovirus") %>% 
  ggplot(aes(x = taxon_genus, y = total_rpk, fill = Condition)) +
  geom_col(position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c("Case" = "#d73027", "Control" = "#4575b4"), labels = c("Case", "Control")) +
  geom_text(aes(label = no_of_peptides), 
            position = position_dodge(width = 0.9), 
            vjust = -0.2, hjust = -0.7, size = 3) +
  geom_text(aes(label = no_of_samples), 
            position = position_dodge(width = 0.9), 
            vjust = -2, hjust = -0.7, size = 3, color = "purple") +
  facet_wrap(~ saccver) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom",
        line = element_blank(),
      axis.text.x = element_blank(), 
      strip.text = element_text(face = "bold")) +
  labs(x = "Enterovirus peptides with significant hits", y = "Total normalised abundance", fill = "")
```

### Stats VIGR 

No significance for VIGR in any fixed effects for any models. However, the total sample size for VIGR is small (42). Furthemore, after stratifying by sex, only ~10 males or females remain in both case and control conditions, which therefore limits the statistical power. 

```{r}
vigr_motif_for_stats <- vigr_motif_results %>% 
  mutate(Nest = str_extract(sample_id, "\\d+")) %>% 
  filter(taxon_genus == "Enterovirus") %>% 
  group_by(saccver, sample_id) %>% #removed condition
  mutate(total_rpk = sum(rpk)) %>% 
  ungroup() %>% 
  select(saccver, Condition, total_rpk, sample_id, taxon_genus, 
         Nest, Age, Sex, HLA_Status, My_name) %>% 
 distinct(saccver, taxon_genus, sample_id, .keep_all = TRUE) %>% 
 mutate(case = ifelse(Condition == "Case", 1, 0),
        log_total_rpk = log(total_rpk + 1),
        across(c(HLA_Status, Sex, Nest, My_name), as.factor))
```

Number of males and females per case or control group:
```{r}
vigr_motif_for_stats %>% 
  distinct(sample_id, .keep_all = TRUE) %>% 
  count(Condition, Sex, name = "count")
```

#### VIGR 3D region

```{r}
vigr_3d_motif_for_stats <- vigr_motif_for_stats %>% filter(saccver == "3D_motif")

vigr_3d_model <- glmmTMB(formula = case ~ log_total_rpk + HLA_Status + Age + Sex + (1|Nest) + (1|My_name),
       family = poisson(link = "log"),
       data = vigr_3d_motif_for_stats)

vigr_3d_model %>% tidy_looksie()

```

#### VIGR VP1 region

```{r}
vigr_vp1_motif_for_stats <- vigr_motif_for_stats %>% filter(saccver == "VP1_motif")

vigr_vp1_model <- glmmTMB(formula = case ~ log_total_rpk + HLA_Status + Age + Sex + (1|Nest) + (1|My_name),
       family = poisson(link = "log"),
       data = vigr_vp1_motif_for_stats)

vigr_vp1_model %>% tidy_looksie()
```

#### VIGR VP1 Sex stratified 

Females VP1 motif
```{r}
vigr_vp1_motif_for_stats_xx <- vigr_vp1_motif_for_stats %>% filter(Sex == "F")

vigr_vp1_xx_model <- glmmTMB(formula = case ~ log_total_rpk + HLA_Status + Age + (1|Nest) + (1|My_name),
       family = poisson(link = "log"),
       data = vigr_vp1_motif_for_stats_xx)

vigr_vp1_xx_model %>% tidy_looksie()
```

Males VP1 motif

```{r}
vigr_vp1_motif_for_stats_xy <- vigr_vp1_motif_for_stats %>% filter(Sex == "M")

vigr_vp1_xy_model <- glmmTMB(formula = case ~ log_total_rpk + HLA_Status + Age + (1|Nest) + (1|My_name),
       family = poisson(link = "log"),
       data = vigr_vp1_motif_for_stats_xy)

vigr_vp1_xy_model %>% tidy_looksie()
```

